A-maze of Natural Stories: Texts are comprehensible using the Maze task

### 2021-08-23 Papers/Paper/manuscript.Rmd

---
title             : "A-maze of Natural Stories: Comprehension and surprisal in the Maze task TODO title"
shorttitle        : "A-maze of Natural Stories"

@@

TODO overall: go through Roger's comments!
TODO fix tense!

@@

TODO top level description

@@

## Reading stories in the Maze task

@@

TODO: question: Does the current order of all Maze then all SPR results make sense? Or would be it better to have Maze-GAM, SPR-GAM, Maze-LM, SPR-LM, etc?

@@

(ref:gam-cap) GAM predictions of reaction time (RT) as a function of either current word surprisal (top) or previous word surprisal (bottom). Density of data is shown along the x-axis. For each of the 4 language models used, there is a linear relationship between current word surprisal and RT (at least when there is enough data). The relationship between previous word surprisal and RT is much flatter.

@@

The GAMs predicted relationship between current and previous word surprisals are shown in Figure \@ref(fig:gam); this is the main effect of either current or previous surprisal for an average length and frequency word, if the other surprisal value is 0 (TODO Not sure this is the correct interpretation! -- https://rdrr.io/cran/tidymv/man/get_gam_predictions.html is the function used for prediction). Note that for each of the models, high-surprisal words are rare, with much of the data for words between 0 and 15 bits of surprisal. All of the models show a roughly linear relationship between current word surprisal and RT, especially in the region with more data. All of the models show a flatter relationship between previous word surprisal and RT. This is a sign of localization as the previous word's surprisal is not affecting RT much. The linear relationship matches that found in SPR and eye-tracking. TODO comment on freq, length effects / non-linearities!

As shown in Table \@ref(tab:pre-error), we found large effects of surprisal and length, but minimal effects of frequency. The lack of frequency effects is somewhat surprising, but consistent with @shainLargescaleStudyEffects2019. Notably the coefficients for the lagged terms are small relative to the effects of surprisal and length of the current word, an indication that spillover is limited and effects are strongly localized.

As a last analysis, we checked which of our surprisal models had the best fit using a nested model comparison shown in Table \@ref(tab:maze-compare). We assess the benefits of adding each model's predictions as a second surprisal source to see which models pick up on information not contained in another model.  GPT-2 provides a lot of additional predictive value over each other model, GRNN provides a lot over 5-gram and TXL and a little complemetary information over GPT-2. TXL provides a lot over 5-gram, and 5-gram provides little over any model. Log likelihood, which is a measure of fit to data, shows  a similar hierarchy, as GPT-2 is better than GRNN is better than TXL is better than 5-gram.
```{r maze-compare}

## Comparison with SPR

@@

@@

As a comparison, we ran the same types of models on the Self-Paced Reading data collected by @futrellNaturalStoriesCorpus2020. As shown in Figure \@ref(fig:spr-gam), there is a roughly linear but fairly flat relationship between RT and surprisal. Note that the y-axis is fairly narrow, and so the predicted effect of changes in surprisal is fairly small. This is confirmed by linear models (see Table \@ref(tab:spr-table)). Surprisal and length effects are evident for the current word, and most models shows surprisal effects from the past word, but we do not see frequency effects. The effects are much smaller than for A-maze, even though this model accounts for spillover effects from more previous words.

Similarly to the model comparison for Maze, we also conducted a model comparison on the SPR data shown in Table \@ref(tab:spr-compare).  Model comparisons show that GPT-2 and 5-gram models contain some value over each other model, which is less clear for TXL and GRNN. In terms of log likelihoods, we find that GPT-2 is better than 5-gram is better than GRNN is better than TXL, although differences are small. The relatively good fit of 5-gram models to SPR data compared with neural models matches results from @huSystematicAssessmentSyntactic2020 and @wilcoxPredictivePowerNeural2020.  This contrasts with the Maze results, where the 5-gram model has the worst fit and does not provide additional predictive value to the other models.

@@

@@

An additional comparison between SPR and Maze is how correlated their reading times are; that is, do the methods pick up on the same effects. If so, we would expect that the sentences that take longer to read in one method also take longer in the other. We calculate the average RT at the sentence level (see Methods for details). The correlation between Maze and SPR is `r round(spr_maze,2)`, compared to `r round(spr_spr,2)` within SPR and `r round(maze_maze,2)` within Maze. See Figure \@ref(fig:scatter) for a visual comparison of overall Maze versus SPR RTs.

TODO mention at sentence level to control for spillover profiles and so single words are too noisy

Add interpretation

@@

We took advantage of the pre-existing SPR corpus on Natural Stories to compare the RT profiles between Maze and SPR. While these are not completely comparably as the subject populations differed, we are able to get a sense of the relationship between SPR and Maze results. Maze and SPR pick up on similar features in words, as shown by the correlations between Maze and SPR, and the fact that surprisal and length are predictive of RTs in both. However, the effect sizes of RT and surprisal are much larger is Maze, and the effects are more strongly localized on a single word. Further comparisons between different processing on the same materials could be useful for identifying how task demands influence language processing [ex. @bartekSearchOnlineLocality2011].

>>>>

---
title             : "A-maze of Natural Stories: Comprehension and surprisal in the Maze task"
shorttitle        : "A-maze of Natural Stories"

@@

@@

@@

## Reading in the Maze

@@

### GAMs

@@

(ref:gam-cap) GAM predictions of Maze reaction time (RT) as a function of either current word surprisal (top) or previous word surprisal (bottom). Density of data is shown along the x-axis. For each of the 4 language models used, there is a linear relationship between current word surprisal and RT (at least when there is enough data). The relationship between previous word surprisal and RT is much flatter.

@@

The GAMs predicted relationship between current and previous word surprisals are shown in Figure \@ref(fig:gam); this is the main effect of either current or previous surprisal for an average length and frequency word, if the other surprisal value is 0 (TODO Not sure this is the correct interpretation! -- https://rdrr.io/cran/tidymv/man/get_gam_predictions.html is the function used for prediction). Note that for each of the models, high-surprisal words are rare, with much of the data for words between 0 and 15 bits of surprisal. All of the models show a roughly linear relationship between current word surprisal and RT, especially in the region with more data. All of the models show a flatter relationship between previous word surprisal and RT. This is a sign of localization as the previous word's surprisal is not affecting RT much.

TODO comment on freq, length effects / non-linearities!

@@

As a comparison, we also ran GAMs on the Self-Paced Reading data collected by @futrellNaturalStoriesCorpus2020. Previous work such as @smithEffectWordPredictability2013 has found positive relationships between RT and the surprisal of earlier words for SPR, so we include predictors from the 3 prior words.  The predicted relationship between surprisals and RT is shown in Figure \@ref(fig:spr-gam); note that the y-axis range is muuch narrower than for Maze. Both current and previous word surprisals have a roughly linear positive relationship to RT. The surprisal of the word two back also has an influence in some models.

Comparing the Maze and SPR smooths, we see that both show a linear relationship, but Maze has much larger and more localized effects of surprisal.

### Effect sizes

The Maze linear model effects are shown in Table \@ref(tab:pre-error).  we found large effects of surprisal and length, but minimal effects of frequency.  The lack of frequency effects is somewhat surprising, but consistent with @shainLargescaleStudyEffects2019. Notably the coefficients for the lagged terms are small relative to the effects of surprisal and length of the current word, an indication that spillover is limited and effects are strongly localized.

@@

We ran similar models for SPR, although to account for known spillover effects, we consider predictors from the current and 3 previous words. Due to issues fitting models, the details of the models differed (see Methods).  The SPR coefficients are shown in Table \@ref(tab:spr-table). Surprisal, length, and frequency effects are all evident for the current word and surprisal and frequency show effects from the previous word as well.

Whereas Maze showed surprisal effects in the 10 to 25 ms/bit range and length effects 15-20 ms/character range, SPR effects are about 1/10 this size. This difference in effect is disproportionate to the overall speed of the methods; Maze is 2-3 times as slow as SPR.

In addition, the relative effects of these different predictors shows somewhat different patterns; in SPR, frequency effects are roughly as big as surprisal effects, but this is not true in Maze. In SPR, length effects exceed surprisal effects, in Maze, this varies depending on the model.

TODO: is there a way to make the tables more readable / highlight the key rows? Maybe stick the full tables in the appendix and show select versions omitting interactions and 2- 3- past words in the main text? Or maybe it's fine and people can read?

### Model comparison

We use 4 different models for surprisal. This provided robustness to the above analyses, but it also allows us to look at which models have the best fit to human Maze and SPR data. For each data type, we assessed each models fit using log likelihood and R-squared. Then we did a nested model comparison, looking at whether a model with multiple surprisal predictors (ex, GRNN and GPT) had a better fit than a model with only one (ex GRNN alone). This tells us whether a model contains predictive value beyond that contained in another model. As shown in Table \@ref(tab:maze-compare), GPT-2 provides a lot of additional predictive value over each other model, GRNN provides a lot over 5-gram and TXL and a little complementary information over GPT-2. TXL provides a lot over 5-gram, and 5-gram provides little over any model. The single-model measures of log likelihood confirm this hierarchy, as GPT-2 is better than GRNN is better than TXL is better than 5-gram.
```{r maze-compare}

We followed the same process for the SPR data with results shown in Table \@ref(tab:spr-compare).  For SPR, GPT-2 and 5-gram models contain some value over each other model, which is less clear for TXL and GRNN. In terms of log likelihoods, we find that GPT-2 is better than 5-gram is better than GRNN is better than TXL, although differences are small. The relatively good fit of 5-gram models to SPR data compared with neural models matches results from @huSystematicAssessmentSyntactic2020 and @wilcoxPredictivePowerNeural2020.  This contrasts with the Maze results, where the 5-gram model had the worst fit and did not provide additional predictive value to the other models.

@@

### Maze - SPR correlation

@@

As a last analysis, we measured how correlated SPR and Maze are in their reading times; that is, did the methods pick up on the same effects. If so, sentences that take longer to read in one method would also take longer in the other. We calculate the average RT at the sentence level because differences in spillover will effect sentence level RT less than RT for individual words (see Methods for details). The correlation between Maze and SPR is `r round(spr_maze,2)`, compared to `r round(spr_spr,2)` within SPR and `r round(maze_maze,2)` within Maze. See Figure \@ref(fig:scatter) for a visual comparison of overall Maze versus SPR RTs.

The fact that SPR data is about as correlated with Maze as with another sample of SPR data provides some confirmation that Maze and SPR are measuring the same thing.

@@

We took advantage of the pre-existing SPR corpus on Natural Stories to compare the RT profiles between Maze and SPR. While these are not completely comparably as the subject populations differed, we are able to get a sense of the relationship between SPR and Maze results. Maze and SPR pick up on similar features in words, as shown by the correlations between Maze and SPR, and the fact that surprisal and length are predictive of RTs in both. However, the effect sizes of RT and surprisal are much larger is Maze, and the effects are more strongly localized on a single word.

Some language models are better or worse on predicting human data; however, we find preliminary evidence that how predictive a language model is of human data may depend on what task was used to collect the RTs, with the 5-gram model relatively better at predicting SPR data. Further comparisons between different processing on the same materials could be useful for identifying how task demands influence language processing [ex. @bartekSearchOnlineLocality2011].
